Beyond Infrastructure: Understanding the Causal Impact of Bike Lane Installation on Urban Cycling Adoption

### 2024-12-02 paper/paper.qmd
final edits to paper, cleaned up scripts structure

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Steven Li
thanks: "Code and data are available at: [https://github.com/stevenli-uoft/Toronto_BikeShare_Causality](https://github.com/stevenli-uoft/Toronto_BikeShare_Causality)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
```

\newpage
# Introduction
Context and motivation paragraph

Estimand paragraph

Results paragraph

Why it's important

Paper structure

# Data {#sec-data}
## Overview
Tools, data sources, and context

## Measurement
- Detailed explanation of how cycling infrastructure changes are measured and recorded
- Discussion of how ridership is measured through the Bike Share system
- Talk about how the data goes from people to paper/data

## Data Cleaning
Overview of data cleaning steps to reach final dataframe. Mention that a much more details data processing methodology is outlined in appendix.

**discuss using relative year for pre/treatment/post periods**

## Outcome Variables
- Monthly ridership counts
- Seasonally adjusted ridership

\vspace*{\fill}
\begin{center}
**Line graph of ridership over relative months, both raw and adjusted riderhsip**
\end{center}
\vspace*{\fill}

## Treatment Variables
- Treatment: Infrastructure changes (with timing)
- Include visualization of treatment and control groups.

\vspace*{\fill}
\begin{center}
**Bar graph of upgrades and installation over time**
\end{center}
\vspace*{\fill}

## Summary Statistics
\vspace*{\fill}
\begin{center}
**Table of treatment vs control group statistics, pre/post treatment**
\end{center}
\vspace*{\fill}

# Model
## Model justification
Why DiD model

## Assumptions
Assumptions on data and definitons

## Model set-up
Mathematical notation of model

## Robustness Checks
### Parallel Trends Test
- Methodology and justification

### Placebo Tests
- Methodology and justification

### Heterogeneity Analysis
- Define subgroups (Types of bikeways)
- Methodology and justification

# Results
## Parallel Trend Analysis
\vspace*{\fill}
\begin{center}
**Visualization of parallel trend test**
\end{center}
\vspace*{\fill}

- Statistics and discussion

## Difference-in-Difference Model Estimates
\vspace*{\fill}
\begin{center}
**Visualization of DiD model**
\end{center}
\vspace*{\fill}

- Primary results
- Discussion of statistical results

## Robustness Checks
### Placebo Test Results
- Present results and discuss implications

### Heterogenous Effects
- Compare effects across bikeway types
- Include figures or tables

# Discussion
## Model Results Interpretation

## Policy Implications and Recommendations

## Limitations and Future Research

>>>>

---
title: "Beyond Infrastructure: Understanding the Causal Impact of Bike Lane Installation on Urban Cycling Adoption"
subtitle: "A Difference-in-Differences Analysis of Toronto's Cycling Network, 2017-2023"
author: 
  - Steven Li
thanks: "Code and data are available at: [https://github.com/stevenli-uoft/Toronto_BikeShare_Causality](https://github.com/stevenli-uoft/Toronto_BikeShare_Causality)."
date: today
date-format: long
abstract: "We examine the causal effect of cycling infrastructure improvements on Bike Share Toronto ridership by analyzing how different types of bikeways influence system usage between 2017-2023. Using a difference-in-differences design with over 7.4 million rides across 1,191 bikeways, we find that infrastructure improvements led to significant increases in ridership, with post-implementation effects reaching 133.1 additional monthly rides per bikeway, translating to approximately 300,000 additional annual rides across all treated segments. Protected lanes demonstrate the strongest impact, generating 381.9 additional monthly rides - more than double the effect of on-road lanes (169.2 rides) and triple that of shared roadways (111.7 rides), while upgrades to existing infrastructure showed more immediate benefits than new installations. Our findings suggest that strategic implementation of protected infrastructure can substantially increase cycling activity, with implications for Toronto's planned addition of 100 kilometers of cycling infrastructure between 2025-2027. These results provide evidence-based guidance for cities working to encourage sustainable transportation through cycling infrastructure development."
format:
  pdf:
    toc: true
include-before: |
  \newpage
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(dplyr)
library(knitr)
library(lubridate)
library(ggplot2)
library(arrow)
library(kableExtra)
library(fixest)
library(modelsummary)

analysis_data <- read_parquet(here::here("data/02-analysis_data/final_df.parquet"))
```

\newpage
# Introduction
Cycling infrastructure plays a crucial role in urban transportation, yet quantifying its impact on cycling adoption remains challenging. While protected bike lanes can reduce cycling injuries by up to 89% [@harris2020lane] and increase ridership by up to 171% on specific routes [@nacto2016], less is known about how different types of infrastructure improvements affect bike share system utilization. This knowledge gap is particularly relevant for Toronto, where rapid expansion of cycling infrastructure has coincided with dramatic growth in the bike share network from 2,854 to over 6,850 stations between 2016 and 2020 [@toronto_bike_share_expansion]. Understanding this relationship is critical as cities increasingly rely on bike share systems to complement public transit and provide sustainable transportation options.

We estimate the causal effect of cycling infrastructure changes on Bike Share Toronto ridership between 2019-2021, examining how protected lanes, on-road lanes, and shared roadways influence system usage. Using a difference-in-differences design, we analyze both new installations and infrastructure upgrades while controlling for broader trends in cycling adoption.

Our analysis reveals substantial increases in bike share ridership following infrastructure improvements. Post-implementation effects reach 133.1 additional monthly rides per bikeway (p < 0.01), translating to approximately 300,000 additional annual rides across all 186 treated segments - particularly notable as these improvements represent only 16% of the 1,191 bikeways in our study. Protected lanes demonstrate the strongest impact, generating 381.9 additional monthly rides (p < 0.001) - more than double the effect of on-road lanes and triple that of shared roadways. These gains occurred despite treated bikeways having higher initial ridership at 209.03 monthly rides, compared to control bikeways 85.59 monthly rides pre-treatment. This suggests infrastructure improvements can substantially increase cycling activity even on popular routes.

These findings carry important implications for urban transportation policy. Toronto's planned addition of 100 kilometers of cycling infrastructure between 2025-2027 [@cyclingNetworkPlan] presents an opportunity to enhance cycling adoption through strategic implementation. Our results challenge conventional assumptions about prioritizing underserved areas, as improvements to high-traffic routes generated significant additional ridership, highlighting the complex trade-offs in infrastructure resource allocation.

The remainder of this paper proceeds as follows. @sec-data describes our data sources and measurement approach. @sec-model outlines our difference-in-differences methodology and identification strategy. @sec-results presents our main results and robustness checks. @sec-discussion discusses the implications of our findings and potential directions for future research. @sec-ideal-survey introduces an idealized survey methodology designed to directly address our research question. The appendices provide supplementary materials, including detailed data processing steps in @sec-data-processing and additional figures and analysis in @sec-additional-figures.

# Data {#sec-data}
## Overview
This study draws on three datasets from the City of Toronto's Open Data Portal [@torontoOpenData]: the Cycling Network (Bikeways) data [@bikeWayData], and both Bike Share Ridership and Stations data [@bikeShareData]. Together, these datasets provide a comprehensive view of Toronto's cycling landscape, capturing infrastructure development from 2001-2023 and ridership patterns from 2017-2023. All data cleaning and analysis was conducted using R [@RCoreTeam2023], with key packages including `tidyverse` [@tidyverse], `dplyr` [@dplyr], and `sf` [@sf] for spatial analysis.

## Measurement and Limitations
### Cycling Infrastructure Measurement
The Cycling Network data captures two distinct types of infrastructure changes recorded by the City's Transportation Services division [@bikeWayData]:

1. **New Installation**: The construction of new bikeway segments that expand the network's reach
2. **Upgrades**: Modifications to existing segments that enhance their safety or functionality

When either change occurs, the location, infrastructure type, and year of modification are documented. A key limitation of this measurement process is its temporal granularity - while the year of each change is recorded, the specific month and day are not, constraining our ability to pinpoint exact implementation dates.

### Ridership and Stations Measurement
Bike Share ridership is measured through an automated system that anonymously records every trip's start and end points, time, and associated station IDs. The measurement process follows these steps:

1. A user unlocks a bike using their membership card, mobile app, or payment kiosks located at each station
2. The system logs the start time and station information
3. Upon return, the end time and station information are recorded
4. This data is automatically transmitted to Bike Share Toronto's database

The Public Bicycle System Company (PBSC) and City of Toronto also maintains detailed Bike Share stations records, documenting each station's location and capacity metrics [@bikeWayData].

### Limitations and Potential Biases
Our analysis faces two primary measurement challenges that could affect causal inference. First, the Bike Share system cannot track actual route choices between endpoints. To address this, we focus on trips originating from stations within 100 meters of bikeways, assuming proximity indicates usage. While this provides a reasonable proxy, it may introduce selection bias by excluding riders who use bikeways despite starting from more distant stations.

Second, our measure of infrastructure changes may suffer from reverse causality bias. The city might prioritize upgrades for routes that already show high cycling traffic, potentially overstating the causal effect of infrastructure improvements on usage. We attempt to mitigate this through our difference-in-differences design and by examining pre-treatment trends, but this limitation should be considered when interpreting our results.

## Data Cleaning
The data cleaning process involved several key steps to create our final analysis panel dataset (detailed methodology available in Appendix @sec-data-processing):

1. Spatial filtering to identify stations proximate to bikeways (within 100 meters)
2. Extract relevant Bike Share trips based on filtered station locations
3. Aggregating total monthly rides for each bikeway, adjusting for seasonality
4. Creating treatment and control groups:

   - **Treatment group**: Bikeways upgraded or constructed between 2019-2021
   - **Control group **: Bikeways unmodified since before 2017

A crucial methodological choice was the use of relative timing rather than calendar years. Due to bikeways having different upgrade or installation years, we align their treatment event year using a relative period. For treated bikeways, this creates a 60-month window: 24 months pre-treatment, 12 months during the treatment year, and 24 months post-treatment. Control bikeways receive randomly assigned pseudo-treatment years (2019, 2020, or 2021) to enable parallel analysis. This approach allows us facilitate direct comparison between treatment and control groups while accounting for the temporal distribution of infrastructure changes.

@tbl-sample-data in Appendix @sec-additional-figures provides a sample of our processed dataset.

## Outcome Variables
Our primary outcome variable is seasonally-adjusted monthly ridership for each bikeway (`monthly_rides_adj`). @fig-avg-monthly-ridership reveals several important patterns in ridership behavior. The raw ridership (black line) displays pronounced seasonal fluctuations, with consistent peaks during summer months and troughs in winter, reflecting Toronto's strong cycling seasonality. After applying seasonal adjustments (blue line), these cyclical patterns are smoothed while preserving the underlying growth trends. Both treatment and control groups show a general upward trend in ridership over the observation period, suggesting broader growth in cycling activity across Toronto's network.

The notable unadjusted ridership activity during the treatment period (relative month 0) reflects both the granularity of our data and real-world implementation patterns. Since we observe only the year of infrastructure changes due to data limitations, all twelve months of the treatment year are assigned to relative month 0. This ridership activity captures both partial-year effects (as changes may occur at different points throughout the year) and continued system usage during implementation.

```{r}
#| label: fig-avg-monthly-ridership
#| fig-cap: "Average ridership by relative months to treatment"
#| echo: false
#| warning: false
#| message: false

ridership_by_month <- analysis_data %>%
  group_by(relative_month, treatment) %>%
  summarise(
    avg_rides = mean(monthly_rides, na.rm = TRUE),
    avg_rides_adj = mean(monthly_rides_adj, na.rm = TRUE)
  ) %>%
  pivot_longer(
    cols = c(avg_rides, avg_rides_adj),
    names_to = "metric",
    values_to = "rides"
  )

ggplot(ridership_by_month, aes(x = relative_month, y = rides, color = metric)) +
  geom_line() +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(
    x = "Months Relative to Treatment",
    y = "Average Monthly Rides",
    color = "Metric"
  ) +
  scale_color_manual(
    values = c("grey10", "blue"),
    labels = c("Raw Rides", "Seasonally Adjusted")
  ) +
  theme_minimal() +
  theme(legend.position = "top",
        panel.grid.major = element_line(color = "gray85", linewidth = 0.35)
        ) +
  facet_wrap(~treatment, scales = "free_y", 
             labeller = labeller(treatment = c("TRUE" = "Treatment", "FALSE" = "Control")))

```

## Treatment Variables
Our main treatment variable (`treatment`) indicates whether a bikeway received an infrastructure change (either an upgrade or new installation) during our study period 2019-2021. @fig-treatment-control-count presents the distribution of bikeways between treatment and control groups, showing a larger pool of control bikeways to enable robust comparison with treated segments.

Appendix @sec-additional-figures provide a further breakdown on types of bikeways installed and upgraded throughout the years. @fig-bikeway_type_constructed_graph in the appendix shows consistent bikeway installations every year after 2010, with protected lanes most commonly installed. @fig-bikeway_type_upgraded_graph presents yearly bikeway upgrades, showing an increase during our treatment period (2019-2021), especially with protected bike lanes.

```{r}
#| label: fig-treatment-control-count
#| fig-cap: "Total Bikeway Count in Control Vs. Treatment Group"
#| echo: false
#| warning: false
#| message: false

installation_counts <- analysis_data %>%
  select(bikeway_id, treatment, construct_year) %>%
  distinct() %>%
  group_by(treatment) %>%
  summarise(count = n()) %>%
  mutate(
    group_label = ifelse(treatment, "Treatment", "Control")
  )

ggplot(installation_counts, 
       aes(x = group_label, y = count, fill = group_label)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(
    values = c("Control" = "#1f77b4", "Treatment" = "#ff7f0e")
  ) +
  labs(
    x = "Group",
    y = "Number of Bikeways",
    fill = "Group"
  ) +
  theme_minimal() +
  theme(legend.position = "top",
        panel.grid.major = element_line(color = "gray85", linewidth = 0.35)
        )
```

To explore treatment effect heterogeneity, we examine two additional dimensions of infrastructure changes:

- **Bikeway Type**:

  > - *Protected Lanes*: Physically separated cycling infrastructure
  > - *On-Road Lanes*: Dedicated but unprotected bike lanes
  > - *Shared Roadways*: Designated shared-use routes

- **Installation Type**:

  > - *Upgrades*: Improvements to existing infrastructure
  > - *New Installations*: Addition of new cycling routes

@fig-infra-type reveals distinct infrastructure patterns between groups. While protected lanes dominate both treatment and control groups, comprising approximately 50% of bikeways in each, the distribution of other types differs notably. The treatment group shows a higher proportion of on-road lanes relative to shared roadways, suggesting a deliberate strategy of upgrading less protected infrastructure. This aligns with Toronto's broader policy goal of enhancing cyclist safety through infrastructure improvements [@toronto_council_agenda].

@fig-treatment-type-total demonstrates a nearly even split in treatment approaches, with 98 new installations and 88 upgrades among the 186 treated bikeways. This balanced distribution enables us to examine whether infrastructure expansion (new installations) versus enhancement (upgrades) generates different effects on ridership patterns. Further categorization into bikeway types, and sub-treatment groups allow us to examine whether certain types of infrastructures and types of changes yield different impacts on ridership patterns.

```{r}
#| label: fig-infra-type
#| fig-cap: "Infrastrucutre Type Distribution between Control and Treatment Group"
#| echo: false
#| warning: false
#| message: false

infra_counts <- analysis_data %>%
  select(bikeway_id, treatment, bikeway_type) %>%
  distinct() %>%
  group_by(treatment, bikeway_type) %>%
  summarise(count = n())

ggplot(infra_counts, aes(x = bikeway_type, y = count, fill = treatment)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    x = "Infrastructure Type",
    y = "Number of Bikeways",
    fill = "Group"
  ) +
  scale_fill_manual(
    values = c("#1f77b4", "#ff7f0e"),
    labels = c("Control", "Treatment")
  ) +
  theme_minimal() +
  theme(legend.position = "top",
        panel.grid.major = element_line(color = "gray85", linewidth = 0.35)
        )

```

```{r}
#| label: fig-treatment-type-total
#| fig-cap: "Number of New-Installations Vs Upgrades in Treatment Group"
#| echo: false
#| warning: false
#| message: false

treatment_type_counts <- analysis_data %>%
  filter(treatment == TRUE) %>%
  select(bikeway_id, sub_treatment_type) %>%
  distinct() %>%
  group_by(sub_treatment_type) %>%
  summarise(count = n()) %>%
  filter(sub_treatment_type != "No Treatment")

ggplot(treatment_type_counts, aes(x = sub_treatment_type, y = count)) +
  geom_bar(stat = "identity", fill = "#ff7f0e") +
  labs(
    x = "Sub-treatment Type",
    y = "Number of Bikeways"
  ) +
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "gray85", linewidth = 0.35))

```

# Model {#sec-model}
## Assumptions
Our model implementation and interpretation rests on two key assumptions. First, we assume Bike Share rides that start within 100 meters of a bikeway indicate usage of that infrastructure, as cyclists typically choose the most direct route to their destination and are unlikely to travel significant distances to access a bikeway. Second, we treat the year of bikeway construction or upgrade as generating a sharp treatment effect, though we account for potential construction overlap by excluding six months of pre-treatment data in our parallel trend test to mitigate anticipatory affects. Our 24-month post-treatment window allows sufficient time to observe behavioral adaptation while minimizing contamination from other infrastructure changes. Additionally, our use of seasonally adjusted ridership helps isolate infrastructure effects from weather-related variation in cycling patterns, as Toronto's distinct seasonal climate significantly influences cycling behavior.

## Model justification
To estimate the causal effect of bikeway infrastructure changes on Bike Share ridership, we employ a difference-in-differences (DiD) approach. This methodology is particularly well-suited for our research question as it allows us to exploit the temporal variation in bikeway upgrades and installations while controlling for unobserved confounders. The staggered implementation of bikeway improvements across Toronto from 2019 to 2021 provides a natural experiment setting where we can compare ridership changes between treated and untreated bikeways before and after infrastructure improvements.

Our analysis is ran in R [@RCoreTeam2023] using the `fixest` package [@fixest], which provides estimation methods for fixed effects models. This package is particularly suitable for our analysis as it handles multiple fixed effects efficiently and provides robust standard errors clustered at the bikeway level.

## Model Setup
Following the framework of @cuborican2024econometrics, let $y_{it}$ represent the seasonally-adjusted monthly Bike Share rides for bikeway $i$ in month $t$. Our main DiD specification is:

$$y_{it} = \alpha_i + \gamma_t + \sum_{p \in \{treatment,post\}} \beta_p (Treatment_i \times Period_{p,it}) + \epsilon_{it}$$

where:

- $\alpha_i$ represents bikeway fixed effects
- $\gamma_t$ represents relative-month fixed effects
- $Treatment_i$ is an indicator equal to 1 if bikeway $i$ received an upgrade or was newly installed during 2019-2021
- $Period_{p,it}$ are indicators for treatment period and post-treatment period (The pre-treatment period serves as the reference category)
- $\epsilon_{it}$ is the error term

The coefficients of interest are $\beta_{treatment}$ and $\beta_{post}$, which capture the average treatment effects during and after the infrastructure changes, respectively.

## Robustness Checks

### Parallel Trends Test
A crucial assumption of the DiD design is that treated and control bikeways would have followed parallel trends in the absence of treatment [@cuborican2024econometrics]. While this counterfactual cannot be directly tested, we can examine pre-treatment trends to assess the plausibility of this assumption. We implement this test by estimating:

$$y_{it} = \alpha_i + \gamma_t + \sum_{k=-24}^{-6} \delta_k (Treatment_i \times 1[t=k]) + \epsilon_{it}$$

where $k$ indexes months relative to treatment. Due to data limitations that only specify the year of infrastructure changes, we conservatively exclude the six months immediately preceding treatment to account for potential construction overlap from the previous year that could affect ridership patterns. This approach helps isolate the true pre-treatment period from any anticipation effects or early construction impacts. The coefficients $\delta_k$ should not be statistically different from zero if the parallel trends assumption holds.

### Placebo Tests
To further validate our results, we conduct a placebo test by randomly reassigning treatment status among bikeways while maintaining the original proportion of treated units. This approach helps ensure our findings are not driven by spurious correlations or systematic differences unrelated to the actual infrastructure changes. The specification mirrors our main DiD model but uses the randomly assigned treatment status:

$$y_{it} = \alpha_i + \gamma_t + \sum_{p \in \{treatment,post\}} \beta_p (PlaceboTreatment_i \times Period_{p,it}) + \epsilon_{it}$$

### Heterogeneity Analysis
We extend our analysis to examine treatment effect heterogeneity across different types of infrastructure changes. This investigation is crucial for understanding which interventions are most effective at increasing ridership. We estimate two separate models:

1. **Bikeway Type Effects**:

$$y_{it} = \alpha_i + \gamma_t + \sum_{p \in \{treatment,post\}} \sum_{j \in J} \beta_{pj} (Treatment_{ij} \times Period_{p,it}) + \epsilon_{it}$$

where $J = \{$Protected Lanes, On-Road Lanes, Shared Roadways$\}$

2. **Installation Type Effects**:

$$y_{it} = \alpha_i + \gamma_t + \sum_{p \in \{treatment,post\}} \sum_{k \in K} \beta_{pk} (Treatment_{ik} \times Period_{p,it}) + \epsilon_{it}$$

where $K = \{$Upgraded, Newly-Installed$\}$

These specifications allow us to identify which types of infrastructure improvements and implementation strategies are most effective at increasing Bike Share utilization. For both heterogeneity analyses, we cluster standard errors at the bikeway level to account for serial correlation within bikeways over time.

# Results {#sec-results}
## Parallel Trends Analysis
A key assumption of our difference-in-differences design is that treated and control bikeways would have followed parallel trends in the absence of treatment. @fig-parallel visualizes this assumption by plotting the seasonally adjusted monthly rides for both treated and control bikeways during the pre-treatment period. While the treatment group consistently shows higher monthly average ridership levels than the control group, the trends between the two groups appear to move in parallel before treatment, particularly in the period from month -20 to month -10. We exclude the six months immediately preceding treatment from this visual analysis due to possible overlap in construction, avoiding potential anticipation effects.

```{r}
#| label: fig-parallel
#| fig-cap: "Parallel Trends Test: Average Monthly Rides in Pre-Treatment Period"
#| echo: false
#| warning: false
#| message: false

parallel_plot <- analysis_data %>%
  filter(relative_month <= -6)

ggplot(parallel_plot, 
       aes(x = relative_month, 
           y = monthly_rides_adj, color = treatment)) +
  stat_summary(geom = "line", fun = mean, size = 1) +
  labs(
    x = "Relative Months (Pre-Treatment)",
    y = "Seasonally Adjusted Monthly Rides",
    color = "Treatment Status"
  )  +
  scale_color_manual(values = c("TRUE" = "#ff7f0e", "FALSE" = "#1f77b4"),
                     labels = c("Control", "Treated")) +
  theme_minimal() +
  theme(legend.position = "top",
        panel.grid.major = element_line(color = "gray85", linewidth = 0.35)
        )

```

The full results of the parallel trends test are provided in @tbl-parallel in the appendix. The coefficients show no clear trend and are generally not statistically different from zero in the pre-treatment period, supporting the parallel trends assumption. While we observe slightly negative coefficients in months -8 and -7 (significant at the 10% and 5% levels respectively), the magnitudes are small relative to our main treatment effects and may reflect normal variation rather than systematic differences.

## Difference-in-Differences Model Estimates
@tbl-main-results presents the results from our main difference-in-differences model. The model estimates that infrastructure improvements led to a significant increase in Bike Share ridership. During the treatment year, we observe a marginally significant increase of 32.4 rides per month (p < 0.10). The effect becomes substantially larger in the post-treatment period, with an estimated increase of 133.1 rides per month (p < 0.01), compared to the pre-treatment period.

The high R-squared value of 0.785 indicates that our model explains a substantial portion of the variation in ridership. The inclusion of bikeway and time fixed effects helps control for unobserved confounders that might affect ridership patterns.

```{r}
#| label: tbl-main-results
#| tbl-cap: "Main Difference-in-Difference Results"
#| echo: false
#| warning: false
#| message: false

main_did <- readRDS(here::here("models/main_did.rds"))

modelsummary(
  list("Main DiD Model" = main_did),
  stars = TRUE,
  gof_map = c("nobs", "r.squared", "adj.r.squared")
)
```

## Robustness Checks

### Placebo Test Results
To verify that our results are not driven by spurious correlations, we conduct a placebo test by randomly reassigning treatment status across bikeways. As shown in @tbl-placebo, when using this random assignment, we find near-zero effects in both the treatment period (4.0 rides) and post-treatment period (0.1 rides), with neither coefficient being statistically significant. These null results, compared to the significant effects found in our main analysis, support the validity of our findings by demonstrating that randomly assigned "fake" treatments do not generate the ridership increases we observe with actual infrastructure improvements.

```{r}
#| label: tbl-placebo
#| tbl-cap: "Placebo Test Results - Random Treatment Assignment"
#| echo: false
#| warning: false
#| message: false

placebo_random_did <- readRDS(here::here("models/placebo_random_did.rds"))

# Randomly assign treatment
set.seed(778)

placebo_random <- analysis_data %>%
  # First get a list of unique bikeway IDs
  group_by(bikeway_id) %>%
  # Take the first row of each group to get treatment status
  slice(1) %>%
  # Randomly shuffle the treatment status across bikeways
  ungroup() %>%
  mutate(placebo_treatment = sample(treatment, n(), replace = FALSE)) %>%
  # Now join this back to the original data
  select(bikeway_id, placebo_treatment) %>%
  right_join(analysis_data, by = "bikeway_id")

modelsummary(
  list("Placebo Test - Random Assignment" = placebo_random_did),
  stars = TRUE,
  gof_map = c("nobs", "r.squared", "adj.r.squared")
)

```

### Heterogeneous Effects
We examine treatment effect heterogeneity across different types of infrastructure and implementation strategies. @tbl-hetero-lane presents the results by bikeway type, with each coefficient representing the change in ridership compared to the pre-treatment period for that specific infrastructure type. Protected lanes show the most dramatic impact, generating an additional 381.9 rides per month post-treatment (p < 0.001) – more than twice the effect of on-road lanes (169.2 rides, p < 0.001) and more than triple the effect of shared roadways (111.7 rides, p < 0.001). The substantially larger effect of protected lanes suggests that perceived safety significantly influences riders' willingness to use Bike Share.

```{r}
#| label: tbl-hetero-lane
#| tbl-cap: "Heterogeneous Treatment Effects by Bikeway Type"
#| echo: false
#| warning: false
#| message: false

did_hetero_lane <- readRDS(here::here("models/did_hetero_lane_type.rds"))

modelsummary(
  list("Heterogeneous Effects - Bikeway Type" = did_hetero_lane),
  stars = TRUE,
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_map = c(
  "period::treatment:treatment_infra::Control" = "Treatment x Control",
  "period::treatment:treatment_infra::Treated-OnRoad" = "Treatment x Treated-OnRoad",
  "period::treatment:treatment_infra::Treated-Protected" = "Treatment x Treated-Protected",
  "period::treatment:treatment_infra::Treated-Shared" = "Treatment x Treated-Shared",
  "period::post:treatment_infra::Control" = "Post x Control",
  "period::post:treatment_infra::Treated-OnRoad" = "Post x Treated-OnRoad",
  "period::post:treatment_infra::Treated-Protected" = "Post x Treated-Protected",
  "period::post:treatment_infra::Treated-Shared" = "Post x Treated-Shared"
  ),
  output = "latex"
)
```

@tbl-hetero-treat compares the effectiveness of upgrading existing infrastructure versus installing new bikeways. Upgrades to existing infrastructure generate larger and more immediate benefits, showing both a significant treatment-year effect (43.8 additional rides, p < 0.10) and a strong post-treatment effect (152.7 additional rides, p < 0.01). In contrast, newly-installed bikeways show no significant effect during the treatment year and a smaller post-treatment increase (115.5 rides, p < 0.10), suggesting that improvements to established cycling routes may be more effective at increasing ridership than creating entirely new routes.

```{r}
#| label: tbl-hetero-treat
#| tbl-cap: "Heterogeneous Treatment Effects by Implementation Strategy"
#| echo: false
#| warning: false
#| message: false

did_hetero_treat_type <- readRDS(here::here("models/did_hetero_treat_type.rds"))

modelsummary(
  list("Heterogeneous Effects - Treatment Type" = did_hetero_treat_type),
  stars = TRUE,
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  coef_map = c(
    "period::treatment:sub_treatment_type::Upgraded" = "Treatment x Upgraded",
    "period::treatment:sub_treatment_type::Newly-Installed" = "Treatment x Newly Installed",
    "period::post:sub_treatment_type::Upgraded" = "Post x Upgraded",
    "period::post:sub_treatment_type::Newly-Installed" = "Post x Newly Installed"
  ),
  output = "latex"
)
```

These heterogeneity analyses reveal important nuances in the effectiveness of different infrastructure strategies. The particularly strong effect of protected lanes suggests that safety and separation from vehicle traffic may be key factors in encouraging Bike Share usage. Additionally, the stronger effects of upgrades compared to new installations may indicate the importance of improving infrastructure in areas where cycling activity already exists.

# Discussion {#sec-discussion}
Our difference-in-differences analysis of Toronto's cycling infrastructure changes between 2019-2021 reveals nuanced insights about how different types of cycling infrastructure improvements influence Bike Share utilization. The results demonstrate that while infrastructure investments broadly increase cycling adoption, the magnitude and timing of these effects vary substantially by implementation approach and infrastructure type.

## Key Empirical Findings and Their Implications
The evolution of treatment effects over time tells an important story about behavior change and infrastructure adoption. During the treatment year, we observe a modest but marginally significant increase of 32.4 rides per month (p < 0.10). This relatively small initial effect likely reflects several factors. First, infrastructure improvements often cause temporary disruption during construction, potentially suppressing immediate usage of bike share in the surrounding areas. Second, the treatment year impacts may be diluted because we only observe the year of changes, not precise implementation dates. Most importantly, the minor initial increase suggests that behavioral adaptation to new infrastructure takes time as users adjust their travel patterns and routine to new routes.

The significant increase in the post-treatment period – to 133.1 additional rides per month (p < 0.01) – reveals the true potential of infrastructure improvements once fully integrated into the transportation network. This impact from pre-treatment to post-treatment periods aligns with research from the University of Colorado Denver showing that "building safe facilities for cyclists is one of the biggest factors in road safety for everyone" [@MARSHALL2019100539]. The magnitude of this change suggests that infrastructure improvements don't just make existing cyclists' trips safer, they also alter the decisions behind mode of transportation for many residents.

To put these effects in perspective, our treatment estimate of 133.1 additional monthly rides per bikeway translates to approximately 24,757 additional monthly rides across all 186 treated segments, or nearly 300,000 additional rides annually. This substantial increase is reflected in the raw data (see @tbl-summary-statistics in Appendix @sec-additional-figures) - treated segments averaged 331.92 monthly rides post-treatment compared to 209.03 pre-treatment, while control segments saw a more modest increase from 85.59 to 107.77 monthly rides over the same period. The magnitude of this effect is particularly notable given that treated segments represent only about 16% of Toronto's total bikeways (186 out of 1,191 total segments) but generated significant ridership gains, suggesting that targeted infrastructure improvements can have outsized impacts on system utilization.

The dramatic difference in baseline ridership between treatment and control groups (209.03 versus 85.59 average monthly rides pre-treatment, as shown in @tbl-summary-statistics) also provides important context for interpreting our heterogeneous effects. This pattern suggests that the city prioritized improvements on already-popular routes, which may partially explain why upgrades to existing infrastructure showed more immediate effects than new installations. However, the fact that treated segments still demonstrated substantial gains in ridership despite starting from a higher baseline suggests that infrastructure improvements can drive meaningful increases in cycling activity even on already-established routes.

This substantial increase in post-treatment ridership may also be partially driven by strategic expansion of the Bike Share network itself. Following infrastructure improvements, Bike Share Toronto can coordinate with the city to place new stations near cycling infrastructure. This complementary expansion of both infrastructure and Bike Share access could create a multiplier effect, where improved infrastructure attracts more stations, which in turn enables more riders to access the enhanced bike lanes. While this interaction between infrastructure and system expansion complicates our causal interpretation, it reflects the real-world synergy between cycling infrastructure and bike share system development.

### Heterogenous Effect Findings
Our heterogeneous effects analysis reveals crucial insights about which interventions most effectively promote cycling. Protected lanes show remarkably strong effects, generating 381.9 additional monthly rides post-treatment (p < 0.001) – more than twice the impact of on-road lanes (169.2 rides, p < 0.001) and more than triple the effect of shared roadways (111.7 rides, p < 0.001). This substantial difference in effectiveness likely stems from:

- **Safety perceptions**: The physical separation from traffic provided by protected lanes addresses the concerns of what @nacto2016 research identifies as the "interested but concerned" majority of potential cyclists.
- **User confidence**: Protected infrastructure may encourage less experienced cyclists to try Bike Share, expanding the user base beyond confident urban cyclists.
- **Urban density patterns**: Protected lanes are more commonly implemented in denser, high-traffic areas like downtown Toronto, where higher population density and concentration of destinations naturally generate more cycling trips. This geographic distribution may partially explain their stronger treatment effects compared to other infrastructure types.

The comparison between upgrades and new installations provides equally important insights for implementation strategy. Upgrades to existing infrastructure generate more monthly ridership (152.7 adjusted rides), compared to newly-installing bike lanes (115.5 adjusted rides) post-treatment. Additionally, during the treatment period, upgrades to existing infrastructure show immediate effect to ridership (43.8 rides, p < 0.10), whereas new installations show no significant treatment-year effect. This pattern suggests that improving existing cycling routes may be more effective than expanding into new areas, particularly for short-term ridership gains. This finding aligns with Toronto's strategic focus during the same period as our study, on upgrading existing infrastructure to "higher-order" facilities, as outlined in the 2019-2021 Cycling Network Plan Update [@toronto_council_agenda].

## Infrastucture Development and Policy Implications
Our findings have substantial implications for cycling infrastructure development strategies. The strong performance of protected lanes, combined with the effectiveness of upgrades, suggests a coupled approach:

- Systematically upgrade existing high-traffic routes to protected infrastructure where feasible
- Strategically expand the network through new protected lanes in high-potential corridors

This approach is particularly relevant given Toronto's commitment to add 100 kilometers of new cycling infrastructure between 2025 and 2027 [@cyclingNetworkPlan]. The significant post-treatment effects we observe – especially for protected lanes – suggest these investments could substantially increase Bike Share ridership if properly implemented. These findings and implications align with historical actions, as in 2020, the City of Toronto worked to expand Bike Share coverage to 20 of the city's 25 wards (Toronto Parking Authority 2020), further suggesting opportunities for coordinated deployment of infrastructure improvements and system expansion.

The timing of effects also has important implications for policy evaluation. The delay between implementation and peak impact suggests that infrastructure evaluations should maintain longer time horizons. Short-term metrics may significantly underestimate the full benefits of cycling investments, particularly for new installations which show no significant treatment-year effects but meaningful long-term impacts. This delayed-benefit pattern aligns with experiences in other cities; for instance, Portland saw a 75% reduction in road fatality rates over a 20-year period of sustained infrastructure investment, highlighting the importance of maintaining long-term commitment to cycling infrastructure development [@MARSHALL2019100539].

## Methodological Strengths and Limitations
Our difference-in-differences design, supported by robust parallel trends in the pre-treatment period and null results in placebo tests, provides strong causal evidence for the effectiveness of cycling infrastructure. The robustness of our results is particularly noteworthy given Toronto's challenging research context, where multiple infrastructure projects and transportation initiatives often overlap temporally and spatially. However, several limitations merit discussion:
